Analysing Suicide Event Patterns Show Trend amongst Crisis Calls

### 2024-01-25 outputs/paper/paper.qmd
final changes

library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)

@@

Toronto's Person In Crisis lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis. The frequency of crises in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents. Over the years, the situation has drastically changed. As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call center handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines. This is quite a large number of calls being received. He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021.

The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction. A variety of event types, including attempt suicide, the person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the data set. The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014,2015. In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in the Toronto neighborhood. To identify significant patterns and trends within the Persons in Crisis calls for help attended data set, several examinations of the interactions between type of events, geographical locations, month, day of the week & time of the day, and event-specific aspects are used. The focus of this paper is to draw the correlation between suicide-related events and the time of the day or day of the week. The analysis of suicide-related events, their frequency, and the geographical area of PIC calls shows a definitive trend among these factors. The analysis aims to provide insightful information that will improve the crisis response system and promote community well-being.

@@

# graph displays frequency of all the different type of events
 ggplot(cleaned_data, aes(x = event_type)) +
 geom_bar(fill = "pink") +
 labs(title = "Event Types and Their Frequencies", x = "Event Type", y = "Frequency", color = "Country") +
 scale_color_brewer(palette = "Set1") +
 theme(legend.position = "bottom")
```

@@

# makes a plot of all the frequency of the suicide related calls in each neighbourhood
ggplot(cleaned_data_suicide, aes(x = factor(hood_158), fill = event_type)) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Event Type by Hood 158", x = "Hood 158", y = "Count") +
  theme_minimal() +
  theme(legend.position = "top",        # Change legend position
        legend.title = element_blank()) +
theme(axis.text.x=element_text(size=0)) # to not show the x-axis as too many numbers
```

@@

@@

time <- cleaned_data %>%
  mutate(Time=as.POSIXct(event_hour, format="%H")) %>%
  ggplot(aes(Time, fill=fct_infreq(event_dow))) +
  geom_density(adjust=0.5, position="stack") +
  labs(x = "Hour", y = "Density", fill= "Day of Week") +
  scale_x_continuous(breaks = seq(0, 24, by = 2)) +
  scale_fill_manual(values=day_colors) +
  theme(axis.text.x=element_text(size=8)) +
  ggtitle("Frequency of Suicide Calls")

time

@@

# creates a correlation graph between time of the day and Suicide calls
# code inspiration from https://stackoverflow.com/questions/66652330/error-ggplot-continuous-x-aesthetic-did-you-forget-aesgroup
cleaned_data_suicide1 |>
    filter(event_type == "Suicide-related") %>%
  group_by(event_hour) %>% # calculating by hours
  summarise(suicide_frequency = n()) %>% # to get the frequency of the suicide calls
  ggplot(aes(x = factor(event_hour), y = suicide_frequency, label = "Calls")) + geom_point() + 
  ggrepel::geom_text_repel() + 
labs(
    title = "Frequency of PIC Calls By The Hour",
    x = "Hour of the Day",
    y = "Frequency of Calls"
  ) +
  theme_classic()

@@

There is also a clear a correlation between the time of the day and day of the week with the frequency of suicide-related PIC calls like [@fig-correlation]. Thursday is the day of the week that receives most of the calls with Sunday being the day that receives the least calls. The hours between 1300 to 1500 receive a high amount of calls compared to the other time of the day which shows a direct relation between the day and time and suicide related calls.

>>>>

library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)
library(reshape2)

@@

Toronto's Person In Crisis(PIC) lines are an invaluable community asset that provides free, anonymous assistance to people experiencing a serious crisis. The frequency of crises in today’s cities, particularly those involving populations at risk (PICs), highlights the need to better understand the dynamics behind such incidents. Over the years, the situation has drastically changed. As stated by Beverly Romeo-Beehler [@resource1], Before the COVID-19 pandemic hit in 2018 and 2019, the call center handled over 1.9 million service calls a year, with slightly under 60% of those calls coming from emergency 9-1-1 lines. This is quite a large number of calls being received. He further adds that calls for service fell to 1.7 million in 2020 and then slightly increased to 1.8 million in 2021.

The Toronto Police Service is in charge of the system, which is built to handle several kinds of crisis calls, each of which calls for a unique reaction. A variety of event types, including attempt suicide, the person in crisis, elopee, jumper, overdose, and threatened suicide, are included in the data set. The geographical location of the calls according to Toronto's hood_158 system has also been recorded in the year 2014 and 2015. In this paper, all the neighborhood (Hood_158) data is used as an estimand to get a better understanding of all the hotspots of the PIC calls in the Toronto neighborhood. To identify significant patterns and trends within the Persons in Crisis calls for help attended data set, several examinations of the interactions between type of events, geographical locations, month, day of the week & time of the day, and event-specific aspects are used. The focus of this paper is to draw the correlation between suicide-related events and the time of the day or day of the week. The analysis of suicide-related events, their frequency, and the geographical area of PIC calls shows a definitive trend among these factors. The analysis aims to provide insightful information that will improve the crisis response system and promote community well-being.

@@

# graph displays frequency of all the different type of events
# More appealing to show through a graph rather than a table, numbers are easier to visualize and put into perspective
 ggplot(cleaned_data, aes(x = event_type)) +
 geom_bar(fill = "pink") +
 labs(title = "Event Types and Their Frequencies", x = "Event Type", y = "Frequency", color = "Country") +
 scale_color_brewer(palette = "Set1") +
 theme(legend.position = "bottom")
```

@@

# makes a plot of all the frequency of the suicide related calls in each neighbourhood
ggplot(cleaned_data_suicide, aes(x = factor(hood_158), fill = event_type)) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Event Type by Hood 158", x = "Hood 158", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none",        # Change legend position
        legend.title = element_blank()) +
theme(axis.text.x=element_text(size=0)) # to not show the x-axis as too many numbers
```

@@

# A heatmap for the data visualization of days and time of PIC suicide calls

# code inspiration from: https://r-graph-gallery.com/215-the-heatmap-function.html

table_data <- table(cleaned_data_suicide1$event_dow, cleaned_data_suicide1$event_hour)
df <- as.data.frame.matrix(table_data)
 # Reorder days of the week
df$Day <- factor(row.names(df), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
# Melt the data for ggplot
melted_df <- melt(df, id.vars = "Day")
# Create a heatmap
ggplot(melted_df, aes(x = variable, y = Day, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Suicide Calls by Day of the Week and Hour of the Day",
       x = "Hour of the Day",
      y = "Day of the Week") +
  theme_minimal() + 
  theme(axis.text.x=element_text(size=6))

@@

# This is another example of the map for visualization, run it if you want a better understanding with colors.

#time <- cleaned_data %>%
#  mutate(Time=as.POSIXct(event_hour, format="%H")) %>%
#  ggplot(aes(Time, fill=fct_infreq(event_dow))) +
#  geom_density(adjust=0.5, position="stack") +
#  labs(x = "Hour", y = "Density", fill= "Day of Week") +
#  scale_x_continuous(breaks = seq(0, 24, by = 5)) +
#  scale_fill_manual(values=day_colors) +
#  theme(axis.text.x=element_text(size=8)) +
#  ggtitle("Frequency of Suicide Calls")

#time

@@

# creates a correlation graph between time of the day and Suicide calls
# code inspiration from https://stackoverflow.com/questions/66652330/error-ggplot-continuous-x-aesthetic-did-you-forget-aesgroup
cleaned_data_suicide1 |>
    filter(event_type == "Suicide-related") %>%
  group_by(event_hour) %>% # calculating by hours
  summarise(suicide_frequency = n()) %>% # to get the frequency of the suicide calls
  ggplot(aes(x = factor(event_hour), y = suicide_frequency)) + geom_point() + 
labs(
    title = "Frequency of PIC Calls By The Hour",
    x = "Hour of the Day",
    y = "Frequency of Calls"
  ) +
  theme_classic()

@@

There is also a clear a correlation between the time of the day and day of the week with the frequency of suicide-related PIC calls like [@fig-correlation]. Thursday is the day of the week that receives most of the calls with Sunday being the day that receives the least calls. The hours between 1500 to 1600 receive a high amount of calls compared to the other time of the day which shows a direct relation between the day and time and suicide related calls.
